No one-size-fits-all solution to clean GBIF

### 2020-06-14 manuscript_peerj.tex
manuscript resubmission

The number of records flagged by individual tests was similar across study groups, suggesting that similar problems might be relevant for collections of plants and animals. Therefore, the same filters can be used across taxonomic groups. Some notable exceptions stress the need to adapt each filter to the taxonomic study group to balance data quality and data availability. The high fraction of records filtered by the ``basis of record filter'' for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups. The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700. The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700. The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and \emph{Oocephalus} might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.

The general agreement between the species richness maps based on raw and filtered data was encouraging, for the use of this data for large-scale biogeographic research (Fig. \ref{fig:speciesrichness}, Appendix 2). The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from a previous study in the coffee family (Maldonado et al. 2015). Records assigned to country or province centroids are often old records, which are geo-referenced at a later point based on vague locality descriptions. These records are at the same time more likely to represent dubious species names, since they might be old synonyms or type specimens of species that have only been collected and described once, which are erroneously increasing species numbers.

We overall consider the effect of the automated filters as positive since they identified the above-mentioned issues and increased the data precision and reduced computational burden (Table 3, Appendix 2). However, in some cases filters failed to remove major issues, often due to incomplete metadata. For instance, for Diogenidae we found at least two records of an extinct species (\emph{Paguristes mexicanus}) which slipped the ``basis of record'' test because they were marked as ``preserved specimen'' rather than ``fossil specimen''. Furthermore, for Entomobryidae we found that for 1,996 records the meta-data on taxonomic rank was ``UNRANKED'' despite all of them being identified to species level, leading to a high fraction of records removed by the ``Identification level'' filter. Additionally automated filters might be overly strict or unsuitable for certain taxa. For instance, in Entomobryidae, 2,004 samples were marked as material samples, and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.

The filters we included in this study address a set of important but relatively easy to identify problems. In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, \url{https://data-blog.gbif.org/post/country-centroids/}, \url{https://data-blog.gbif.org/post/country-centroids/}). While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system allowing users to provide expert assessments, i.e.~a meta-annotation of records being challenged (and why). Such a system would not need to change the original data and could include multiple levels to account for differing opinions.

As a next step for automated filtering, it is possible to detect additional problematic records with the help of external species distribution information, which is available for some taxa for example from www.iucn.org (vertebrates) or \url{https://wcsp.science.kew.org/} (selected seed plant families). Furthermore, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups and often are contradictory within individual taxa.

@@

This study enrolled participants of the workshop ``Biodiversity data: from field to yield'' led by Alice Calvente, Fernanda Carvalho, Alexander Zizka, and Alexandre Antonelli through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI. We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for financial support (edital 02/2016 - internacionalização). AZ is funded by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv. AA is supported by the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew. FS was financed by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior - Brasil (CAPES) - Finance Code 001 and Fundação de Amparo à Pesquisa do estado de São Paulo (FAPESP) (FAPESP, process 2015/20215-7).

@@

\begin{itemize}
\item
  Appendix 1 - Absolute number of flagged records per taxonomic group and test
\item
  Appendix 2 - Taxon specific richness maps and comments
\item
  Appendix 3 - Full results of the conservation assessment
\end{itemize}

@@

\caption{\label{tab:tabletaxa}The study groups and their taxonomy. This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders.}
\centering
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{3cm}}
\toprule
 & Taxon & Taxon rank & Realm & Common name & 'Phylum' & Family\\
\midrule
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/diogenidae.png} & Diogenidae & Family & Marine & Hermit crabs & Arthropoda & Diogenidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/entomobryidae.png} & Entomobryidae & Family & Terrestrial & Springtails & Arthropoda & Entomobryidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/neanuridae.png} & Neanuridae & Family & Terrestrial & Springtails & Arthropoda & Neanuridae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tityus.png} & \em{Tityus} & Genus & Terrestrial & Scorpions & Arthropoda & Buthidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/arhynchobatidae.png} & Arhynchobatidae & Family & Marine & Skates & Chordata & Arhynchobatidae\\
\addlinespace
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/dipsadidae.png} & Dipsadidae & Family & Terrestrial & Snakes & Chordata & Dipsadidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/harengula.png} & \em{Harengula} & Genus & Marine & Herrings & Chordata & Clupeidae\\
\hline
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/thozetella.png} & \em{Thozetella} & Genus & Terrestrial & Sac fungi & Ascomycota & Chaetosphaeriaceae\\
\hline
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/conchocarpus.png} & \em{Conchocarpus} & Genus & Terrestrial & NA & Angiosperms & Rutaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/gaylussacia.png} & \em{Gaylussacia} & Genus & Terrestrial & Huckleberries & Angiosperms & Ericaceae\\
\addlinespace
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/harpalyce.png} & \em{Harpalyce} & Genus & Terrestrial & NA & Angiosperms & Fabaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/iridaceae.png} & Iridaceae & Family & Terrestrial & NA & Angiosperms & Iridaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/lepismium.png} & \em{Lepismium} & Genus & Terrestrial & Cacti & Angiosperms & Cactaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/oocephalus.png} & \em{Oocephalus} & Genus & Terrestrial & NA & Angiosperms & Lamiaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/pilosocereus.png} & \em{Pilosocereus} & Genus & Terrestrial & Cacti & Angiosperms & Cactaceae\\
\addlinespace
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/prosthechea.png} & \em{Prosthechea} & Genus & Terrestrial & Orchids & Angiosperms & Orchidaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tillandsia.png} & \em{Tillandsia} & Genus & Terrestrial & Bromeliads & Angiosperms & Bromeliaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tocoyena.png} & \em{Tocoyena} & Genus & Terrestrial & NA & Angiosperms & Rubiaceae\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:tabletests}The automated filters used in this study.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{9cm}}
\toprule
Test & Type & Basis & Rationale\\
\midrule
\rowcolor{gray!6}  Biodiversity institutions & Error & Gazetteer-based & Records may have coordinates at the location of biodiversity institutions, e.g. because they were erroneously entered with the physical location of the specimen or because they represent individuals from captivity or horticulture\\
Equal lat/lon & Error & Gazetteer-based & Coordinates with equal latitude and longitude are usually indicative of data entry errors\\
\rowcolor{gray!6}  Sea & Error & Gazetteer-based & Coordinates from terrestrial organisms in the sea are usually indicative of data entry errors, e.g. swapped latitude and longitude\\
Zeros & Error & Gazetteer-based & Coordinates with plain zeros are often indicative of data entry errors\\
\rowcolor{gray!6}  Capitals & Unfit & Gazetteer-based & Records may be assigned to the coordinates of country capitals based on a vague locality description\\
\addlinespace
Duplicates & Unfit & Gazetteer-based & Duplicated records may add unnecessary computational burden, in particular for large scale biodiversity analyses and distribution modelling for many species\\
\rowcolor{gray!6}  Political centroids & Unfit & Gazetteer-based & Records may be assigned to the coordinates of the centroids of political entities based on a vague locality description\\
Urban areas & Unfit & Gazetteer-based & Records from urban areas are not necessarily errors, but often represent imprecise records automatically geo-referenced from vague locality descriptions or old records from different land-use types\\
\rowcolor{gray!6}  Basis of record & Unfit & Meta-data & Records might be unsuitable or unreliable for certain analyses dependent on their source, e.g. "fossil" or "unknown"\\
Collection year & Unfit & Meta-data & Coordinates from old records are more likely to be imprecise or erroneous coordinates since they are derived from  geo-referencing based on the locality description. This is more problematic for older records, since names or borders of places may change\\
\addlinespace
\rowcolor{gray!6}  Coordinate precision & Unfit & Meta-data & Records may be unsuitable for a study if their precision is lower than the study analysis scale\\
Identification level & Unfit & Meta-data & Records may be unsuitable if they are not identified to species level.\\
\rowcolor{gray!6}  Individual count & Unfit & Meta-data & Records may be unsuitable if the number of recorded individuals is 0 (record of absence) or if the count is too high, as this is often related to records from barcoding or indicative of data entry problems.\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:tablecoords}The impact of automated filtering on occurrence records for 18 Neotropical taxa downloaded from www.gbif.org. From column four onwards the numbers show the percentage of records flagged by the respective test. Only tests that flagged at least 0.1\% of the records in any group are shown. Individual records can be flagged by multiple tests, therefore the sum of percentage from all tests can supersede the total percentage.}

@@

\caption{\label{tab:unnamed-chunk-3}Conservation assessment for 11 Neotropical taxa of plants and animals based on three datasets. IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: Preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: Preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset. Only taxa with at least on species evaluated by IUCN shown.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{IUCN} & \multicolumn{3}{c}{GBIF Raw} & \multicolumn{5}{c}{GBIF Filtered} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-12}
Taxon & n taxa & Evaluated [\%] & Threatened [\%] & n taxa & Threatened [\%] & Match with IUCN [\%] & n taxa & Threatened [\%] & Match with IUCN [\%] & EOO change compared to raw [\%] & AOO change compared to raw [\%]\\
\midrule
Arhynchobatidae & 37 & 51.3 & 17.9 & 39 & 35.9 & 45.0 & 39 & 41.0 & 40.0 & -32.7 & -18.5\\
Dipsadidae & 520 & 68.0 & 8.8 & 638 & 58.3 & 63.0 & 598 & 59.9 & 61.2 & -2.3 & -15.6\\
\em{Harengula} & 4 & 100.0 & 0.0 & 4 & 0.0 & 100.0 & 4 & 0.0 & 100.0 & -38.0 & -36.9\\
\hline
\em{Conchocarpus} & 4 & 8.7 & 0.0 & 46 & 63.0 & 100.0 & 45 & 62.2 & 100.0 & -15.3 & -7.1\\
\em{Gaylussacia} & 2 & 3.3 & 0.0 & 61 & 59.0 & 50.0 & 58 & 60.3 & 50.0 & -22.5 & -8.6\\
\addlinespace
\em{Harpalyce} & 3 & 15.0 & 5.0 & 20 & 65.0 & 66.7 & 17 & 58.8 & 50.0 & -18.4 & -16.5\\
Iridaceae & 13 & 2.3 & 0.2 & 531 & 64.4 & 50.0 & 466 & 62.9 & 62.5 & -18.2 & -12.3\\
\em{Lepismium} & 6 & 100.0 & 0.0 & 6 & 16.7 & 83.3 & 6 & 16.7 & 83.3 & -33.9 & -7.9\\
\em{Pilosocereus} & 41 & 80.9 & 19.1 & 47 & 55.3 & 73.7 & 46 & 56.5 & 71.1 & -8.5 & -5.8\\
\em{Tillandsia} & 54 & 11.6 & 6.0 & 464 & 61.4 & 85.2 & 453 & 62.7 & 83.3 & -13.7 & -9.9\\
\addlinespace
\em{Tocoyena} & 3 & 13.6 & 4.5 & 22 & 31.8 & 66.7 & 21 & 38.1 & 66.7 & -23.0 & -9.5\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

>>>>

The number of records flagged by individual tests was similar across study groups, suggesting that similar problems might be relevant for collections of plants and animals. Therefore, the same filters can be used across taxonomic groups. Some notable exceptions stress the need to adapt each filter to the taxonomic study group to balance data quality and data availability. The high fraction of records filtered by the ``basis of record'' filter for \emph{Tityus}, Dipsadidae, \emph{Prosthechea} and \emph{Tillandsia}, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups. The high fraction of records flagged by the ``collection year'' filter for Dispadidae was caused by a high collection effort in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700. The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institution's code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700. The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and \emph{Oocephalus} might be related to horticulture, since at least some species in those groups are commonly cultivated as ornamentals. This was supported by the detailed examination of the data for Iridaceae, which showed that after filtering 1605 records from 69 exotic species remained in the dataset, stressing the importance to address these species in certain taxonomic groups.

The general agreement between the species richness maps based on raw and filtered data was encouraging, in terms of the use of this data for large-scale biogeographic research (Fig. \ref{fig:speciesrichness}, Appendix 2). The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from a previous study in the coffee family (Maldonado et al. 2015). Records assigned to country or province centroids are often old records, which are geo-referenced at a later point based on vague locality descriptions. These records are at the same time more likely to represent dubious species names, since they might be old synonyms or type specimens of species that have only been collected and described once, which are erroneously increasing species numbers.

Overall, we consider the effect of the automated filters as positive since they identified the above-mentioned issues and increased the data precision and reduced computational burden (Table 3, Appendix 2). However, in some cases filters failed to remove major issues, often due to incomplete metadata. For instance, for Diogenidae we found at least two records of an extinct species (\emph{Paguristes mexicanus}) which slipped the ``basis of record'' test because they were marked as ``preserved specimen'' rather than ``fossil specimen''. Furthermore, for Entomobryidae we found that for 1,996 records the meta-data on taxonomic rank was ``UNRANKED'' despite all of them being identified to species level, leading to a high fraction of records removed by the ``Identification level'' filter. Additionally automated filters might be overly strict or unsuitable for certain taxa. For instance, in Entomobryidae, 2,004 samples were marked as material samples and therefore removed by our global filter retaining only specimen and observation data, which in this case was overly strict.

The filters we included in this study address a set of important but relatively easy to identify problems. In fact, the internal quality control of GBIF does flag some of the problems we tested for (i.e., zero coordinates, equal lat/lon) while others might be implemented in the near future (country centroids, \url{https://data-blog.gbif.org/post/country-centroids/}, \url{https://data-blog.gbif.org/post/country-centroids/}). While this internal quality is very helpful, we see a huge potential to overcome issues with data quality in a user-feedback system that allows users to provide expert assessments, i.e.~a meta-annotation of records being challenged (and why). Such a system would not need to change the original data and could include multiple levels to account for differing opinions.

As a next step for automated filtering, it is possible to detect additional problematic records with the help of external species distribution information, which is available for some taxa, for example from www.iucn.org (vertebrates) or \url{https://wcsp.science.kew.org/} (selected seed plant families). Furthermore, while in this study we focused on meta-data and geographic filtering, taxonomic cleaning---the resolution of synonymies and identification of accepted names---is another important part of data curation, but depends on taxon-specific taxonomic backbones and synonymy lists which are not readily available for many groups and often are contradictory within individual taxa.

@@

We thank GBIF and all data collectors and contributors for their excellent work. We thank Town Peterson, Roderic Page and one anonymous reviewer for the helpful comments on an earlier version of this manuscript. This study enrolled participants of the workshop ``Biodiversity data: from field to yield'' led by Alice Calvente, Fernanda Carvalho, Alexander Zizka, and Alexandre Antonelli through the Programa de Pós Graduação em Sistemática e Evolução of the Universidade Federal do Rio Grande do Norte (UFRN) and promoted by the 6th Conference on Comparative Biology of Monocotyledons - Monocots VI. We thank the Pró-reitoria de Pesquisa and the Pró-reitoria de Pós-graduação of UFRN for financial support (edital 02/2016 - internacionalização). AZ is funded by iDiv via the German Research Foundation (DFG FZT 118), specifically through sDiv, the Synthesis Centre of iDiv. AA is supported by the Swedish Research Council, the Knut and Alice Wallenberg Foundation, the Swedish Foundation for Strategic Research and the Royal Botanic Gardens, Kew. FS was financed by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior - Brasil (CAPES) - Finance Code 001 and Fundação de Amparo à Pesquisa do estado de São Paulo (FAPESP) (FAPESP, process 2015/20215-7).

@@

\begin{itemize}
\item
  Appendix 1 - Absolute number of flagged records per taxonomic group and test
\item
  Appendix 2 - Taxon-specific richness maps and comments
\item
  Appendix 3 - Full results of the conservation assessment
\end{itemize}

@@

\caption{\label{tab:tabletaxa}The study groups and their taxonomy. This study includes three marine and 15 terrestrial taxa, seven of them animals, one group of fungi and ten plants, belonging to 16 different orders. The outlines illustrate the broad taxonomic group. Pictures from www.phylopic.org. Phylopic credits: Diogenidae (Ekaterina Kopeykina (vectorized by T. Michael Keesey), Creative Commons Attribution-NonCommercial-ShareAlike 3.0; Entomobryidae and Neanuridae: Birgit Lang, Creative Commons Attribution 3.0; Tityus: Gareth Monger, Creative Commons Attribution 3.0; \textit{Gaylussacia}, \textit{Lepismium}, \textit{Pilosocereus}: T. Michael Keesey (vectorization) and Nadiatalent (photography), Creative Commons Attribution-ShareAlike 3.0; \textit{Iridaceae} adapted from Goldblatt and Manning (2008); \textit{Oocephalus}: Alexander Schmidt-Lebhuhn, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; \textit{Tillandsia}: Maija Karala, Creative Commons Attribution-NonCommercial-ShareAlike 3.0; all others in the public domain.}
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{1.5cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{3cm}}
 & Taxon & Taxon rank & Realm & Common name & 'Phylum' & Family\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/diogenidae.png} & Diogenidae & Family & Marine & Hermit crabs & Arthropoda & Diogenidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/entomobryidae.png} & Entomobryidae & Family & Terrestrial & Springtails & Arthropoda & Entomobryidae\\
\includegraphics[valign=T, scale=0.2, raise= 2mm]{input/phylopics/converted/neanuridae.png} & Neanuridae & Family & Terrestrial & Springtails & Arthropoda & Neanuridae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tityus.png} & \em{Tityus} & Genus & Terrestrial & Scorpions & Arthropoda & Buthidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/arhynchobatidae.png} & Arhynchobatidae & Family & Marine & Skates & Chordata & Arhynchobatidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/dipsadidae.png} & Dipsadidae & Family & Terrestrial & Snakes & Chordata & Dipsadidae\\
\includegraphics[valign=T, scale=0.18, raise= 2mm]{input/phylopics/converted/harengula.png} & \em{Harengula} & Genus & Marine & Herrings & Chordata & Clupeidae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/thozetella.png} & \em{Thozetella} & Genus & Terrestrial & Sac fungi & Ascomycota & Chaetosphaeriaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/conchocarpus.png} & \em{Conchocarpus} & Genus & Terrestrial & NA & Angiosperms & Rutaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/gaylussacia.png} & \em{Gaylussacia} & Genus & Terrestrial & Huckleberries & Angiosperms & Ericaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/harpalyce.png} & \em{Harpalyce} & Genus & Terrestrial & NA & Angiosperms & Fabaceae\\
\includegraphics[valign=T, scale=0.075, raise= 2mm]{input/phylopics/converted/iridaceae.png} & Iridaceae & Family & Terrestrial & NA & Angiosperms & Iridaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/lepismium.png} & \em{Lepismium} & Genus & Terrestrial & Cacti & Angiosperms & Cactaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/oocephalus.png} & \em{Oocephalus} & Genus & Terrestrial & NA & Angiosperms & Lamiaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/pilosocereus.png} & \em{Pilosocereus} & Genus & Terrestrial & Cacti & Angiosperms & Cactaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/prosthechea.png} & \em{Prosthechea} & Genus & Terrestrial & Orchids & Angiosperms & Orchidaceae\\
\includegraphics[valign=T, scale=0.25, raise= 2mm]{input/phylopics/converted/tillandsia.png} & \em{Tillandsia} & Genus & Terrestrial & Bromeliads & Angiosperms & Bromeliaceae\\
\includegraphics[valign=T, scale=0.2, raise= 2mm]{input/phylopics/converted/tocoyena.png} & \em{Tocoyena} & Genus & Terrestrial & NA & Angiosperms & Rubiaceae\\

@@

\caption{\label{tab:tabletests}The automated filters used in this study.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{9cm}}
\toprule
Test & Type & Basis & Rationale\\
\midrule
\rowcolor{gray!6}  Biodiversity institutions & Error & Gazetteer-based & Records may have coordinates at the location of biodiversity institutions, e.g. because they were erroneously entered with the physical location of the specimen or because they represent individuals from captivity or horticulture\\
Equal lat/lon & Error & Gazetteer-based & Coordinates with equal latitude and longitude are usually indicative of data entry errors\\
\rowcolor{gray!6}  Sea & Error & Gazetteer-based & Coordinates from terrestrial organisms in the sea are usually indicative of data entry errors, e.g. swapped latitude and longitude\\
Zeros & Error & Gazetteer-based & Coordinates with plain zeros are often indicative of data entry errors\\
\rowcolor{gray!6}  Capitals & Unfit & Gazetteer-based & Records may be assigned to the coordinates of country capitals based on a vague locality description\\
\addlinespace
Duplicates & Unfit & Gazetteer-based & Duplicated records may add unnecessary computational burden, in particular for large scale biodiversity analyses and distribution modelling for many species\\
\rowcolor{gray!6}  Political centroids & Unfit & Gazetteer-based & Records may be assigned to the coordinates of the centroids of political entities based on a vague locality description\\
Urban areas & Unfit & Gazetteer-based & Records from urban areas are not necessarily errors, but often represent imprecise records automatically geo-referenced from vague locality descriptions or old records from different land-use types\\
\rowcolor{gray!6}  Basis of record & Unfit & Meta-data & Records might be unsuitable or unreliable for certain analyses dependent on their source, e.g. 'fossil' or 'unknown'\\
Collection year & Unfit & Meta-data & Coordinates from old records are more likely to be imprecise or erroneous coordinates since they are derived from  geo-referencing based on the locality description. This is more problematic for older records, since names or borders of places may change\\
\addlinespace
\rowcolor{gray!6}  Coordinate precision & Unfit & Meta-data & Records may be unsuitable for a study if their precision is lower than the study analysis scale\\
Identification level & Unfit & Meta-data & Records may be unsuitable if they are not identified to species level.\\
\rowcolor{gray!6}  Individual count & Unfit & Meta-data & Records may be unsuitable if the number of recorded individuals is 0 (record of absence) or if the count is too high, as this is often related to records from barcoding or indicative of data entry problems.\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:tablecoords}The impact of automated filtering on occurrence records for 18 Neotropical taxa downloaded from www.gbif.org. From column four onwards the numbers show the percentage of records flagged by the respective test. Only tests that flagged at least 0.1\% of the records in any group are shown. Individual records can be flagged by multiple tests, therefore the sum of percentages from all tests can supersede the total percentage.}

@@

\caption{\label{tab:unnamed-chunk-3}Conservation assessment for 11 Neotropical taxa of plants and animals based on three datasets. IUCN: global red list assessment obtained from www.iucn.org; GBIF Raw: preliminary conservation assessment based on IUCN Criterion B using ConR and the raw dataset from GBIF; GBIF filtered: preliminary conservation assessment based on IUCN Criterion B using ConR and the filtered dataset. Only taxa with at least one species evaluated by IUCN shown.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.2cm}>{\raggedleft\arraybackslash}p{1.5cm}>{\raggedleft\arraybackslash}p{1.5cm}}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{IUCN} & \multicolumn{3}{c}{GBIF Raw} & \multicolumn{5}{c}{GBIF Filtered} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-12}
Taxon & n taxa & Evaluated [\%] & Threatened [\%] & n taxa & Threatened [\%] & Match with IUCN [\%] & n taxa & Threatened [\%] & Match with IUCN [\%] & EOO change compared to raw [\%] & AOO change compared to raw [\%]\\
\midrule
Arhynchobatidae & 37 & 51.3 & 17.9 & 39 & 35.9 & 45.0 & 39 & 41.0 & 40.0 & -32.7 & -18.5\\
Dipsadidae & 520 & 68.0 & 8.8 & 638 & 58.3 & 63.0 & 598 & 59.9 & 61.2 & -2.3 & -15.6\\
\em{Harengula} & 4 & 100.0 & 0.0 & 4 & 0.0 & 100.0 & 4 & 0.0 & 100.0 & -38.0 & -36.9\\
\hline
\em{Conchocarpus} & 4 & 8.7 & 0.0 & 46 & 63.0 & 100.0 & 45 & 62.2 & 100.0 & -15.3 & -7.1\\
\em{Gaylussacia} & 2 & 3.3 & 0.0 & 61 & 59.0 & 50.0 & 58 & 60.3 & 50.0 & -22.5 & -8.6\\
\addlinespace
\em{Harpalyce} & 3 & 15.0 & 5.0 & 20 & 65.0 & 66.7 & 17 & 58.8 & 50.0 & -18.4 & -16.5\\
Iridaceae & 13 & 2.3 & 0.2 & 531 & 64.4 & 50.0 & 466 & 62.9 & 62.5 & -18.2 & -12.3\\
\em{Lepismium} & 6 & 100.0 & 0.0 & 6 & 16.7 & 83.3 & 6 & 16.7 & 83.3 & -33.9 & -7.9\\
\em{Pilosocereus} & 41 & 80.9 & 19.1 & 47 & 55.3 & 73.7 & 46 & 56.5 & 71.1 & -8.5 & -5.8\\
\em{Tillandsia} & 54 & 11.6 & 6.0 & 464 & 61.4 & 85.2 & 453 & 62.7 & 83.3 & -13.7 & -9.9\\
\addlinespace
\em{Tocoyena} & 3 & 13.6 & 4.5 & 22 & 31.8 & 66.7 & 21 & 38.1 & 66.7 & -23.0 & -9.5\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}
